---
title: "Exponential Smoothing"
author: "Simon U., Michael Y., Ben H."
date: "March X, 2020"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()

# Helper to plot dummy time series models for illustration purposes
# x: Numeric vector
# x_hat: Numeric vector of forecast values
# ...: Commands to pass to plot
plot_forecast <- function(y, y_hat, ...) {
    n_y <- length(y)
    n_yhat <- length(y_hat)
    plot(1:(n_y+1), c(y, y_hat[1]), type='l', xlim=c(0, n_y+n_yhat), ylab='y, y_hat', xlab='t', ...)
    lines((n_y+1):(n_y+n_yhat-1), y_hat[2:n_yhat], type='l', col='red')
}
```

# Introduction

# What is exponential smoothing?

Forecasting future observations using weighted averages of past observations, with the weights decaying exponentially as observations recede further into the past

* Presumably, more recent data points are more predictive than older points




# Exponential Smoothing Models

1. Naive---all weight is given to the last observation

2. Average---each past observation is given equal weight

3. Exponential weighted average---Recent observations get higher weight, older observations less weight

4. Holt linear---Same as 3, but accounts for time series with trend

5. Holt-Winters---Same as 4, but also accounts for time series with seasonality

6. State space---



# $ES_{1}$: Naive model

* The naive forecasting model can be thought of us exponential smoothing
* Where 100 percent of weight is given to the last observation:

```{r}
forecast_naive <- function(y, h) {
    n <- length(y)
    y_hat <- rep(y[n], h)
    return( y_hat )
}
```



# $ES_{1}$: Naive model: Example

```{r, fig.height=5}
y <- c(1, 4, 5, 2, 3, 6, 8)
y_hat <- forecast_naive(y, h=7)

plot_forecast(y, y_hat)
```



# $ES_{2}$: Average model

* All future values are forecast as the average of the observed data
* Equivalent to to exponential smoothing where each observation is given equal weight

```{r}
forecast_avg <- function(y, h) {
    y_hat <- rep(mean(y), h)
    return( y_hat )
}
```



# $ES_{2}$: Average model: Example

```{r, fig.height=5}
y_hat <- forecast_avg(y, h=7)
plot_forecast(y, y_hat)
```



# $ES_{3}$: Simple exponential smoothing

* More sophisticated model would given recent observations more weight, and decreasing weight for past observations
* Parameter $\alpha$ controls smoothing; can be optimized

$$\hat{y}_T = \alpha y_T + (1 - \alpha) \hat{y}_{T-1}$$

* $\hat{y}_T \equiv$ predicted value of $y$ at time $t$
* $\alpha \equiv$ user-chosen smoothing parameter, $0 \leq \alpha \leq$
    - 1 means no smoothing, 0 means tons of smoothing
* $\hat{y}_{T-1} \equiv$ predicted value of $y$ at immediately previous period $t - 1$



# $ES_{3}$: Simple exponential smoothing: Problem

__Problem__: What is $\hat{y}_{T-1}$ when $t = 1$?

$$\hat{y}_T = \alpha y_T + (1 - \alpha) \hat{y}_{T-1}$$



* Textbook: Component form, the level $\ell_0$

* Set for mean or median of the time series

# 

```{r}
forecast_simple <- function(y, h, alpha=0.1) {
    n <- length(y)
    l_0 <- mean(y)
    
    l <- c()
    for (i in 1:(n+1)) {
        if (i == 1) {
            l_i <- alpha * y[i] + (1 - alpha) * l_0
        } else {
            l_i <- alpha * y[i] + (1 - alpha) * l[i-1]
        }
        l <- append(l_i, l)
    }
    
    y_hat <- l[n+1]
    return( rep(y_hat, h) )
}
```

# $ES_{3}$: Simple exponential smoothing: Example

```{r}
y_hat <- forecast_simple(y, h=7, alpha=0.25)
plot_forecast(y, y_hat)
```



# $ES_{3}$: Weighted average: Optimizing $\alpha^*$

BEN: Not sure if we want to keep this??



# Interlude

* The previous method is effective for time series without trend or seasonality

* But what if your time series has trend?



# $ES_{4}$: Holt Linear Trend Model

Appropriate for time series that can be described with

$$y_t = \beta_0 + \beta_1 t + \epsilon_t$$

where $\beta_1$ quantifies the trend




# $ES_{4}$: Holt Linear Trend Model: Linear Time Series

* We can explicitly convert our previous time series to a trended one using this formula:

* Previous time series `y = ` (`r y`)

```{r}
beta_0 <- 0
beta_1 <- 1.5
t_ <- 1:7
( y_1 <- beta_0 + y + beta_1*t_ )
```



# $ES_{4}$: Holt Linear Trend Model: Linear Time Series

```{r, echo=FALSE}
plot(t_, y, type='o', ylim=c(0, 20))
lines(t_, y_1, type='o', col='red')
arrows(x0=5, y0=4, x1 = 5, y1 = 9, length = 0.2, angle = 30, col='red', lty=2)
```


# Holt's linear trend + damped

See http://www.real-statistics.com/time-series-analysis/basic-time-series-forecasting/holt-linear-trend/ for excel formula implementation





# Holt-Winters method + multiplicative + taxonomy







# ETS modeling (Innovations state space models)







# Conclusion

# Blah blah


